DP problems of interest

### 2012-09-29 docs/roadmap/problems.tex
More ideas, reworking of first problem (need to do the same for the 3 last).

% LaTeX Template for a short article
\documentclass[11pt]{article}
\usepackage{amssymb,amsmath,amsthm,hyperref,verbatim,pict2e,graphicx,marvosym}
\hypersetup{colorlinks,citecolor=black,filecolor=black,linkcolor=black,urlcolor=black}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage[english]{babel}
 \pagestyle{empty} % remove page numbers

@@

@@

\subsection{Problems classification}
In the literature dynamic programming problems (DP) are classified according to two criteria:

\textbf{Monadic / polyadic}\ul
\item Monadic: on the right hand-side of the recurrence formula, only one term appears. For instance, Smith-Waterman with constant penalty is monadic
	\[M_{(i,j)}=\max\left\{\begin{array}{l} 0 \\ M_{(i-1,j-1)}+{\rm cost}(S(i),T(j))\\ M_{(i-1,j)}-d\\ M_{(i,j-1)}-d \end{array}\right. \]
\item \textbf{Polyadic:} when multiple terms of the recurrence occur in the right and-side of the recurrence formula. For instance Fibonacci is polyadic: \[F(n) = F(n-1) + F(n-2)\]
\ule

\textbf{Serial/non-serial} \ul
\item \textbf{Serial:} when the solution depends only of a fixed number of immediately previous solutions (i.e. neighbor cells). For instance Fibonacci is serial.
\item \textbf{Non-serial:} when the solution of a level depends of an arbitrary number of previous solutions. Typically Smith-Waterman with arbitrary gap penalty and Nussinov are non-serial:
	\[M_{(i,j)}=\max\left\{\begin{array}{l} ... \\ M_{(i,j-1)}\\ \max\limits_{i<k<j} [ M_{(i,k)}+M_{(k+1,j)} ] \end{array}\right. \]
\ule

\subsection{Simplifications}
\subsubsection{Calculus}
In some special case, it is possible to transform a non-serial problem into a serial problem, if we can embed the non-serial term into an additional aggregation matrix. For example, we can transform
	\[M_{(i,j)}=\max\left\{\begin{array}{l} \max\limits_{k<i} M_{(k,j)}
	\\ \sum\limits_{k<i, l<j}M_{(k,l)} \end{array}\right.
	\implies M_{(i,j)}=\max\left\{\begin{array}{l} C_{(k,j)} \\ A_{(i-1,j-1)} \end{array}\right.\]
Where $C$ is a matrix that stores the maximum along the column and $A$ is a matrix that stores the sum of the array of the previous elements. They can be easily computed with the additional recurrence.
	\[\begin{array}{rcl} C_{(i,j)}&=&\max(C_{(i-1,j)}, M_{(i,j)}) \\
	A_{(i,j)}&=&A_{(i-1,j)}+A_{(i,j-1)}-A_{(i-1,j-1)}+M_{(i,j)}\end{array}\]

This simplification avoids the non-serial dependencies at the cost of an additional storage at the wavefront.

\subsubsection{Precomputations}
When a calculus computation is impossible, it might be worth to interleave a computation phase that will aggregate some of the results that are necessary to the computation block. For instance, for Nussinov's $\max_{i<k<j} [ M_{(i,k)}+M_{(k+1,j)} ]$ we can precompute it for all rows and columns of the block, and for all elements that are not part of the block, and pass these partial results together at the block launch.

On GPU, this could be done by interleaving a new kernel for this specific purpose, on FPGA, this could be done by preparing the maximums in another memory area whose pointer will later be passed to the co-processor.

% ----------------------------------------------------------------------------------------------------------------------------------------
\section{Problems of interest}
We describe problems structures: inputs, cost matrices, backtracking matrix. They all have an alphabet (that hopefully corresponds to a maximal bit-size) and matrixes dimensions are specified by number of indices and their number of elements is usually the same as the input. Unless otherwise specified, number are all unsigned integers. We describe the problem processing in terms of both initialization and recurrences.

% ----------------------------------------------
\subsection{Smith-Waterman (simple)}
Problem: matching two strings $S$, $T$.

Alphabets:\ul
\item Input: $\Sigma(S)=\Sigma(T)=\{a,c,g,t\}$.
\item Cost matrix: $\Sigma(M) = {\rm integers} \{0..n\}, n=\max({\rm cost}) \cdot \min({\rm length}(S),{\rm length}(T))$
\item Backtrack matrix: $\Sigma(B)=\{stop,W,N,NW\}$
\ule

@@

% ----------------------------------------------
\subsection{Smith-Waterman (with gap extension at different cost)}
Problem: matching two strings $S$, $T$.

Alphabets:\ul
\item Input: $\Sigma(S)=\Sigma(T)=\{a,c,g,t\}$.
\item Cost matrix: $\Sigma(M) = {\rm integers} \{0..n\}, n=\max({\rm cost}) \cdot \min({\rm length}(S),{\rm length}(T))$
\item Backtrack matrix: $\Sigma(B)=\{stop,W,N,NW\}$
\ule

Initialization:\ul
\item Cost matrix: $M_{(i,0)}=M_{(0,j)}=0$.
\item Gap opening matrix: $E_{(i,0)}= 0, 0 \le i \le {\rm length}(S)$
\item Gap extending matrix: $E_{(0,j)}= 0, 0 \le j \le {\rm length}(T)$
\item Backtrack matrix: $B_{(i,0)}=B_{(0,j)}=stop$.
\ule

Recurrence for the cost matrix:
\[M_{(i,j)}=\max\left\{\begin{array}{l|l}
	0 & stop\\
	M_{(i-1,j-1)}+{\rm cost}(S(i),T(j)) & NW\\
	E_{(i,j)} & N\\
	F_{(i,j)} & W
\end{array}\right\}=B_{(i,j)} \]

Recurrence for the gap opening/extending matrices:
\[E_{(i,j)}=\max\left\{\begin{array}{l|l}
	M_{(i, j-1)} - \alpha & NW\\
	E_{(i,j-1)} - \beta & N\\
\end{array}\right\}=B_{(i,j)} \]

\[F_{(i,j)}=\max\left\{\begin{array}{l|l}
	M_{(i-1, j)} - \alpha & NW\\
	E_{(i-1,j)} - \beta & N\\
\end{array}\right\}=B_{(i,j)} \]

@@

% ----------------------------------------------
\subsection{Convex polygon triangulation}
Problem: triangulating a polygon with least cost overall for added edges

@@

% ----------------------------------------------
\newpage
\subsection{Matrix chain multiplication}
Problem: find an optimal parenthesizing of the multiplication of $n$ matrices $A_i$ of dimension $r_i \times c_i$.

@@

>>>>

% LaTeX Template for a short article
\documentclass[11pt]{article}
\usepackage{amssymb,amsmath,amsthm,hyperref,verbatim,pict2e,graphicx,marvosym,array,booktabs}
\hypersetup{colorlinks,citecolor=black,filecolor=black,linkcolor=black,urlcolor=black}
\usepackage[usenames]{xcolor} % color names ,dvipsnames,svgnames,table
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage[english]{babel}
\pagestyle{empty} % remove page numbers

@@

% Recurrence visualisation
\newcommand\Cd[3][0,-1]{\put(#2){\put(.5,.5){\circle*{.3}}\put(.5,.5){\linethickness{1.5pt}\vector(#1){#3}}}} % dependency [dx,dy]{x,y}{len}
\def\Cg#1{\put(#1){\color{lightgray}\put(0,0){\polygon*(0,0)(0,1)(1,1)(1,0)}}} % grayed cell (not to store
\def\Cz#1{\put(#1){\put(0,.35){\parbox{1\unitlength}{\centering\bf 0}}}} % zero-init cell
\def\Cm{\put(6.5,1.5){\circle*{.4}}\multiput(0,0)(1,0){9}{\line(0,1){8}}\multiput(0,0)(0,1){9}{\line(1,0){8}}} % matrix base

@@

\subsection{Problems classification}
In the literature, dynamic programming problems (DP) are classified according to two criteria:

\textbf{Monadic/polyadic}\ul
\item \textbf{Monadic:} on the right hand-side of the recurrence formula, only one term appears. For instance, Smith-Waterman with constant penalty is monadic
	\[M_{(i,j)}=\max\left\{\begin{array}{l} 0 \\ M_{(i-1,j-1)}+{\rm cost}(S(i),T(j))\\ M_{(i-1,j)}-d\\ M_{(i,j-1)}-d \end{array}\right. \]
\item \textbf{Polyadic:} when multiple terms of the recurrence occur in the right and-side of the recurrence formula. For instance Fibonacci is polyadic: \[F(n) = F(n-1) + F(n-2)\]
\ule

\textbf{Serial/non-serial} \ul
\item \textbf{Serial:} when the solution depends only of a fixed number of immediately previous solutions (i.e. neighbor cells). For instance Fibonacci is serial (it accesses only 2 cells backward).
\item \textbf{Non-serial:} when the solution depends of an arbitrary number of previous solutions. Typically Smith-Waterman with arbitrary gap penalty and Nussinov are non-serial:
	\[M_{(i,j)}=\max\left\{\begin{array}{l} ... \\ M_{(i,j-1)}\\ \max\limits_{i<k<j} [ M_{(i,k)}+M_{(k+1,j)} ] \end{array}\right. \]
\ule

\subsection{Simplifications}
\subsubsection{Calculus}
In some special case, it is possible to transform a non-serial problem into a serial problem, if we can embed the non-serial term into an additional aggregation matrix. For example:
	\[M_{(i,j)}=\max\left\{\begin{array}{l} \max\limits_{k<i} M_{(k,j)}
	\\ \sum\limits_{k<i, l<j}M_{(k,l)} \end{array}\right.
	\implies M_{(i,j)}=\max\left\{\begin{array}{l} C_{(k,j)} \\ A_{(i-1,j-1)} \end{array}\right.\]
Where the matrix $C$ stores the maximum along the column and matrix $A$ stores the sum of the array of the previous elements. Both can be easily computed with an additional recurrence:
	\[\begin{array}{rcl} C_{(i,j)}&=&\max(C_{(i-1,j)}, M_{(i,j)}) \\
	A_{(i,j)}&=&A_{(i-1,j)}+A_{(i,j-1)}-A_{(i-1,j-1)}+M_{(i,j)}\end{array}\]

This simplification avoids the non-serial dependencies at the cost of extra storage in the wavefront, unfortunately, it might be applicable only for some special cases.

\subsubsection{Precomputations}
When a calculus transformation is impossible, it might be worth to interleave a computation phase that will aggregate some of the results that are necessary to the computation block. For instance, for Nussinov term $\max_{i<k<j} [ M_{(i,k)}+M_{(k+1,j)} ]$, we can precompute it over all rows and columns of the block, and for all elements that are not part of the block, then pass these partial results together at the block launch to finish the computation.

On GPU, this could be done by interleaving a new kernel for this specific purpose, on FPGA, this could be done by preparing the maximums in another memory area whose pointer will later be passed to the co-processor. We may notice that since this phase is at the same time necessary for both architecture, and can be run independently, we can both execute them concurrently and mix between architectures (use CUDA for pre-computation and FPGA for actual tile computation for instance).

% ----------------------------------------------------------------------------------------------------------------------------------------
\section{Problems of interest}
We describe problems structures: inputs, cost matrices and backtracking matrix. These all have an alphabet (that must be bounded in terms of bit-size). Unless otherwise specified, we adopt the following conventions:\ul
\item Matrices dimensions are implicitly specified by number of indices and their number of elements is usually the same as the input length.
\item Number are all unsigned integers
\item Problem dimension is $m,n$ (or $n$) indices $i,j$ ranges are respectively $0\le i<m$, $0\le j<n$.
\item Unless otherwise specified, the recurrence applies to all non-initialized matrix elements.
\ule
We describe the problem processing in terms of both initialization and recurrences.

% ----------------------------------------------
\newpage
\subsection{Smith-Waterman (simple)}
Problem: matching two strings $S$, $T$ with $|S|=m, |T|=n$.

Matrices: $M_{m \times n}, B_{m \times n}$

Alphabets:\ul
\item Input: $\Sigma(S)=\Sigma(T)=\{a,c,g,t\}$.
\item Cost matrix: $\Sigma(M) = [0..z], z=\max({\rm cost(\_)}) \cdot \min(m,n)$
\item Backtrack matrix: $\Sigma(B)=\{stop,W,N,NW\}$
\ule

@@

Backtracking: starts from the cell $M_{(m,j)} \cup M_{(i,n)} $

Visualisation:
\begin{center}\setlength{\unitlength}{.6cm}\begin{picture}(8,9)
	\put(-.5,7.5){S}\put(-.35,7.4){\linethickness{1pt}\vector(0,-1){2}}
	\put(.2,8.2){T}\put(.8,8.4){\linethickness{1pt}\vector(1,0){2}}
	\Cz{0,0}\Cz{0,1}\Cz{0,2}\Cz{0,3}\Cz{0,4}\Cz{0,5}\Cz{0,6}\Cz{0,7}
	\Cz{1,7}\Cz{2,7}\Cz{3,7}\Cz{4,7}\Cz{5,7}\Cz{6,7}\Cz{7,7}
	\Cd{6,2}{0.8}
	\Cd[1,0]{5,1}{0.8}
	\Cd[1,-1]{5,2}{0.8}
\Cm\end{picture}\end{center}

Optimizations\ul
\item In serial (monadic) problems we can avoid building the matrix $M$ by only maintaining the 3 last diagonals in memory (one for the diagonal element, one for horizontal/vertical, and one being built). This construction extends easily to polyadic problems where we need to maintain $k+2$ diagonals in memory where $k$ is the maximum backward lookup.
\ule

% ----------------------------------------------
\newpage
\subsection{Smith-Waterman (with gap extension at different cost)}
Problem: matching two strings $S$, $T$ with $|S|=m, |T|=n$.

Matrices: $M_{m \times n}, E_{m \times n}, F_{m \times n}, B_{m \times n}$

Alphabets:\ul
\item Input: $\Sigma(S)=\Sigma(T)=\{a,c,g,t\}$.
\item Cost matrices: $\Sigma(M) = \Sigma(E) = \Sigma(F) = [0..z], z=\max({\rm cost(\_)}) \cdot \min(m,n)$
\item Backtrack matrix: $\Sigma(B)=\{stop,W,N,NW\}$
\ule

Initialization:\ul
\item No gap cost matrix: $M_{(i,0)}=M_{(0,j)}=0$.
\item T-gap extension cost matrix: $E_{(i,0)}= 0$ \textit{<<eat S chars only>>}
\item S-gap extension cost matrix: $F_{(0,j)}= 0$
\item Backtrack matrix: $B_{(i,0)}=B_{(0,j)}=stop$.
\ule

Recurrence for the cost matrices:
\[\begin{array}{rcl}
M_{(i,j)}&=&\max\left\{\begin{array}{l|l}
	0 & stop\\
	M_{(i-1,j-1)}+{\rm cost}(S(i),T(j)) & NW\\
	E_{(i,j)} & N\\
	F_{(i,j)} & W
\end{array}\right\}=B_{(i,j)}\\
\\
E_{(i,j)}&=&\max\left\{\begin{array}{l|l}
	M_{(i, j-1)} - \alpha & NW\\
	E_{(i,j-1)} - \beta & N\\
\end{array}\right\}=B_{(i,j)}\\
\\
F_{(i,j)}&=&\max\left\{\begin{array}{l|l}
	M_{(i-1,j)} - \alpha & NW\\
	F_{(i-1,j)} - \beta & W\\
\end{array}\right\}=B_{(i,j)}
\end{array}\]

{\color{red} XXX: Are we sure we do not need to maintain separate backtrack tables for the gap extensions?}

@@

{\color{red} XXX: how to simplify this relation, can we move the max in the wavefront?}

Recurrence visualisation:

{\color{red} XXX: todo}

Optimizations:

{\color{red} XXX: todo}

% ----------------------------------------------
\newpage
\subsection{Convex polygon triangulation}
Problem: triangulating a polygon with least cost overall for added edges

@@

% ----------------------------------------------
\newpage
\subsection{Matrix chain multiplication}
Problem: find an optimal parenthesizing of the multiplication of $n$ matrices $A_i$ of dimension $r_i \times c_i$ \textit{(r=rows, c=columns)}.

@@

%%\newcolumntype{C}[1]{>{\centering\let\newline\\\arraybackslash}m{#1}}
%\newcolumntype{C}{@{\hspace{7pt}}c@{\hspace{7pt}}}
%\def\mnl{\rule{0pt}{2.6ex}\rule[-1.2ex]{0pt}{0pt} \\ \hline}
%$\begin{array}{|C|C|C|C|C|C|} \hline
%0 & 0 & 0 & 0 & 0 & 0 \mnl
%0 &  &  &  &  & \mnl
%0 & M_{23}  &  &  &  & \mnl
%0 &  & \sum  &  &  & \mnl
%0 &  &  &  &  & \mnl
%0 &  &  &  &  & \mnl
%\end{array}$
